### 2014-04-01 ms_probaweb.tex

Ecology textbooks define the structure of an ecological community as
\emph{the pool of species occupying a given location at a given time, and
the way these species interact} (ref to Morin?). It is therefore surprizing
that ecological research on the variation of community structure has focused
on species turnover (refs), almost entirely neglecting variation in the way
species interact (refs).

\begin{itemize}
\item Network structure do vary in space in time. \\

\item We don't know yet to what extent interactions are varying with the
environment. \\

\item No theory to explain and interpret the meaning of network variation in space.
Current interpretation fo species turnover involves the effect of the
environment and stochasticity. \\

\item Objective: Propose a theoretical framework to understand and predict the
spatial and temporal variation in network structure.\\

\end{itemize}

@@

We will refer to the left term as the metaweb. It is a conditional probability,
representing the probability that an interaction occurs if species $i$ and $j$
are co-occurring. The right term is the probability of observing the two species
co-occurring at location $y$.

% The following paragraph
% (1) is too long -- I tried to cut some of it, up to half of it
% (2) comes at an awkward moment -- this is the key issue we try to solve, it should be in the introduction

The metaweb concept is making its way through the network litterature even
though it has never been formally and technically defined.
% Yes, but no. All of the people that ever wrote the word "metaweb" are authors on this paper.
It is usually conceived as a network of interactions among species
that could potentially co-occur, and represented by a binary matrix,
\emph{i.e.} a probabilistic matrix in which the probability of observing
an interaction is either 0 or 1. As the metaweb is usually constructed by
aggregating observations, it assigns a probability of 0 to the interaction
of species that never co-occurred, and is unable to factor-out the effect of
co-occurrence. The main problem with that approach is that the co-occurrence
of rare species is extremely unlikely and thus most often appear as an absence
of interactions in the metaweb. This approach is inappropriate because the
observed co-occurrence will have a strong signature on the evaluation of
interactions. A rarefaction analysis previously shown that interactions
accumulate with the addition of networks at a slower rate than species
richness \parencite{Poisot2012d}. It indicates that it is harder to have a
direct evaluation of interactions from observeration than it is to evaluate
species richness. If built from the observation of interactions, then the
only way to fill a metaweb is by running cafeteria experiments between all
pairs of species. Otherwise, the metaweb should be inferred using traits
and phylogenetic information. Most of the published metawebs are therefore
incomplete because of their sensitivity to sampling heterogeneity. We will
come back to the issue of evaluating the metaweb in the sections Applications.

There are many variants of the metaweb representing different hypotheses about
the origin of temporal and spatial variation in network structure (see the
explicit formulations at Table 1). First, the interaction could be 
deterministic instead of probabilistic. In other words, $P(L_{ijy}) = 1$ if
$X_{ijy} = 1$, and 0 otherwise. This representation of the metaweb is the one
mostly used so far, as soon as the species are found together they are assumed
to interact. It is also the only way to represent interactions when there is not
enough information available to evaluate the interaction probability. It should be a
reasonnable approximation when the sampling and inferrence scales are large
enough and that the only variation of networks considered arises from species
distribution.

@@

Finally, the co-occurrence probability itself could be dependent on ecological
interactions. Direct pairwise interactions such as competition, facilitation and
predation have long been studied for their impact on co-distribution. Second and
higher order interactions (e.g. trophic cascade) could also impact
co-occurrence. There is however currently no general theory on the expected
co-occurrence in complex ecological networks. For instance, we do not know how
far co-occurrence is not-random when going along the chain of indirect
interactions. Berlow(2009) shown previously that almost only first and second
order interactions do matter in ecological networks, but we don't know for
co-distribution. We neither know what is the sensitiviy to species richness: do
interactions tend to buffer each other? Generalizing knowledge aquired by the
study of small community modiles will require future research.

@@

Part of the problem surrounding the definition of the niche has been
clarified with the distinction between Eltonian and Grinnellian definitions
\parencite{Chase2003}. The Grinnellian dimension of
the niche is the effect of the environment on the demography of a species,
while the Eltonian dimension is the effect of a species on its environment
\emph{sensu lato}.  % SNOB!
The Grinnellian niche is the most intuitive one
to apply and is the conceptual backbone of species distribution models. The
Eltonian niche is well known by community ecologists, but is trickier to
turn into predictive models \parencite{Devictor2010a}. Nonetheless, the
development of the niche model of food web structure (Williams2000) and
its parameterization (Williams2010; Gravel2013) made it more operational,
although it has yet to be applied to more than trophic interactions.

\textcite{Chase2003} attempted this representation in their definition:
\textit{[The niche is] the joint description of the environmental conditions
that allow a species to satisfy its minimum requirements so that the birth rate
of a local population is equal or greater than its death rate along with the
set of per capita effects of that species on these environmental conditions.}
Their representation merges zero-net growth figuring the Grinnelian niche
(when do the population persist) with impact vectors figuring the Eltonian
niche (what is the population impact). While this representation has been very
influential in community ecology at the local scale, it remains unpracticable
%%% WHY?
at the biogeographical one. The absence of any mathematical representation
of the niche that could easily be fit to ecological data perhaps explain why
biogeographers are still struggling to develop species distribution models
taking into account ecological interactions.

@@

In this section we provide an analysis illustrating the framework with an
empirical dataset of host-parasitoid networks. Data come from the study of
Tylianakis(2007) on the impacts of habitat modifications to the network
structure. The data consists of 48 networks with 4090 recorded interactions. The
advantage of replicated host-parasitoid networks is that usually every
interaction is observed, not inferred from a stationary metaweb. It thus allows
to evaluate interaction probability and to factor out the effect of
co-occurrence. Five habitats were sampled along a gradient of habitat
modification: forest, abandonned coffee agroforest, coffee agroforest, pasture
and rice culture. The metaweb consists of 9 parasitoids and kleptoparasites
(Hymenoptera: Eulophidae, Ichneumonidae, Leucospidae, Megachilidae and
Chrysididae; Dyptera: Bombyliidae) of 33 species of bees and wasps (Hymenoptera:
Apidae, Megachilidae, Mutilidae, Pompilidae, Sphecidae, Vespidae). The metaweb
is illustrated at Fig. 2, along with an example of one iteration of the metaweb.

Tylianakis (2007) investigated if habitat modification affects the structure of
these networks They found a significant impact of the habitat on their
structure, despite little variation in species richness. Increasing habitat
modification led to a higher parasitoid to host species ratio and a paraistoid
were also more specialized, thus impacting considerably vulnerability. A closer
inspection of the networks revealed that intensive agricultural systems were
dominated by a strong interaction and a specialization of the most abundanc
parasitoid. Although the discussion made clear that both the turnover in species
composition and the interaction probability changed with habitat modification,
it was not possible to partition these components.

We developped a R package (REF) to fit alternative formulations of the metaweb
and the co-occurrence matrix along an environmental gradient and run it to
re-interpret the data of Tylianaks (2007). The package provides a general
interface facilitating the development of different species and link
distribution models. It is also built to facilitate the interaction iwth the
Mangal database of ecological interactions (REF). The first step consists of
fitting a probablistic model from the observation of a pairwise interaction
(binary) and the environment (could be categorical or continuous) from the
subset of the data where the two species are co-occurring. In other words, it
fits the equation $P(L_{ijy}|X_{iy},X{jy},E_y)$ to the data where $X_{iy} = 1$
and $X_{iy} = 1$. Logistic regression was used and is currently programmed, but
alternative models could be used as well. The second steps consists of fitting a
a probabilistic model for co-occurrence over the whole dataset,
$P(X_{iy},X{jy}|E_y)$, independently of the observation of an interaction. The
two probabilities are then multiplied to obtain the probability of observing an
interaction (Eq. 2). We used this probability to compute the likelihood of each
observation ($\zeta(\theta|D) = P(L_{ijy},X_{iy},X_{jy})$ if $L_{ijy}=1$ and
$\zeta(\theta|D) = 1 - P(L_{ijy},X_{iy},X_{jy})$ otherwise). We then after
compare the models by their AIC.

We considered the gradient of habitat modification as a ordered categorical
variable and compared XX models (results are summarized at Table 2). Not
surprisingly the best model takes into account the effect of the environment on
both the metaweb and co-occurrence. What is most interesting are the comparisons
to the best model. First, we find that using a constant metaweb has a dramatic
impact on the fit of the model to the data (the AIC drops from X for model 1 to
X for model 2), indicating a strong effect of the environment on pairwise
interactions. Secondly, we find that the deterministic metaweb is the worst
model (model 3, AIC = ). This result indicate that the traditional appraoch to
consider that species interact as soon as they co-occur is definitely wrong.
Thirdly, we also find that using a constant co-occurrence does have a
significant impact on the model (the AIC drops to X, model 4), indicating there
is a non-random change in community composition with habitat modification. Taken
together, these two results better explain why network structured changed with
habitat modification, even though here we only used binary information about the
network structure. Another interesting result is that considering a neutral
co-occurrence did not impact much the fit of the modeel. The AIC drops to XX
with model 6, indicating that considering indepdenent SDMs yields similar
networks over this environmental gradient. This means that for this particular
dataset, ecological interactions does not have a strong impact on species
distribution since; a strong dependence of parasitoids to the host for instance
would have a occurrence probability higher than expected by chance, while a
repulsion would have had the opposite.

@@

It is also possible to obtain for each pairwise interaction an estimate of the
uncertainty. Not surprisingly, the confidence interval is usually very high for
the estimation of a probability with a very small sample size. The standard
error on the evaluation of the interaction probability is provided along with
the metaweb at Fig. 3. It reveals that the uncertainty is very high for most
interactions, even if 48 networks were sampled. Such an approach could be used
to detect wich pairwise interaction requires additional sampling in order to
reduce the uncertainty to a manageable level.

>>>>

Ecology textbooks define the structure of an ecological community as
\emph{the pool of species occupying a given location at a given time, and
the way these species interact} (ref to Morin?). It is therefore surprising
that ecological research on the variation of community structure has focused
on species turnover (refs), almost entirely neglecting variation in the way
species interact (refs).

Network structure vary in space and time, metaweb as a large scale
synthesis. We don't know yet the mechanisms of variation

The metaweb concept is making its way through the network literature even
though it has never been formally and technically defined.
It is usually conceived as a network of interactions among species that
could potentially co-occur, and represented by a binary matrix, \emph{i.e.}
a probabilistic matrix in which the probability of observing an interaction
is either 0 or 1. As the metaweb is usually constructed by aggregating
observations, it assigns a probability of 0 to the interaction of species that
never co-occurred, and is unable to factor-out the effect of co-occurrence. The
main problem with that approach is that the co-occurrence of rare species is
extremely unlikely and thus most often appear as an absence of interactions
in the metaweb. A rarefaction analysis previously shown that interactions
accumulate with the addition of networks at a slower rate than species richness
\parencite{Poisot2012d} -- this is because interactions cannot accumulate
faster than the rate of species observations, and ultimately the rate of
co-occurrence observations. The only way to build a metaweb \emph{only}
based on observational data is to compensate for the inflation of 0 due
to lack of co-occurrence through manipulative experiments (REF). As this is
impracticable for all but the smallest organisms, predictive probabilistic
methods are needed.

No theory to explain and interpret the meaning of network variation in space.
Current interpretation of species turnover involves the effect of the
environment and stochasticity.

Objective: Propose a theoretical framework to understand and predict the
spatial and temporal variation in network structure.

@@

We will refer to the left term as the metaweb. It is a conditional
probability, representing the probability that an interaction occurs if
species $i$ and $j$ are co-occurring. The right term is the probability
of observing the two species co-occurring at location $y$. There are many
variants of the metaweb representing different hypotheses about the origin
of temporal and spatial variation in network structure (see the explicit
formulations at Table 1). First, the interaction could be deterministic
instead of probabilistic. In other words, $P(L_{ijy}) = 1$ if $X_{ijy} =
1$, and 0 otherwise. This representation of the metaweb is the one mostly
used so far, as soon as the species are found together they are assumed to
interact. It is also the only way to represent interactions when there is
not enough information available to evaluate the interaction probability. It
should be a reasonable approximation when the sampling and inference scales
are large enough and that the only variation of networks considered arises
from species distribution.

@@

Finally, the co-occurrence probability itself could be dependent on ecological
interactions. Direct pairwise interactions such as competition, facilitation and
predation have long been studied for their impact on co-distribution. Second and
higher order interactions (e.g. trophic cascade) could also impact
co-occurrence. There is however currently no general theory on the expected
co-occurrence in complex ecological networks. For instance, we do not know how
far co-occurrence is not-random when going along the chain of indirect
interactions. Berlow (2009) shown previously that almost only first and second
order interactions do matter in ecological networks, but we don't know for
co-distribution. We neither know what is the sensitivity to species richness: do
interactions tend to buffer each other? Generalizing knowledge acquired by the
study of small community modules will require future research.

@@

Part of the problem surrounding the definition of the niche has been
clarified with the distinction between Eltonian and Grinnellian definitions
\parencite{Chase2003}. The Grinnellian dimension of the niche is the
effect of the environment on the demography of a species, while the Eltonian
dimension is the effect of a species on its environment \emph{sensu lato}. The
Grinnellian niche is the most intuitive one to apply and is the conceptual
backbone of species distribution models. The Eltonian niche is well known
by community ecologists, but is trickier to turn into predictive models
\parencite{Devictor2010a}. Nonetheless, the development of the niche model
of food web structure (Williams2000) and its parameterization (Williams2010;
Gravel2013) made it more operational, although it has yet to be applied to
more than trophic interactions.

\textcite{Chase2003} attempted this representation in their definition:
\textit{[The niche is] the joint description of the environmental conditions
that allow a species to satisfy its minimum requirements so that the birth rate
of a local population is equal or greater than its death rate along with the
set of per capita effects of that species on these environmental conditions.}
Their representation merges zero-net growth figuring the Grinnelian niche
(when do the population persist) with impact vectors figuring the Eltonian
niche (what is the population impact). While this representation has been very
influential in community ecology at the local scale, it remains impracticable
%%% WHY?
at the biogeographical one. The absence of any mathematical representation
of the niche that could easily be fit to ecological data perhaps explain why
biogeographers are still struggling to develop species distribution models
taking into account ecological interactions.

@@

In this section we provide an analysis illustrating the framework with an
empirical dataset of host-parasitoid networks. Data come from the study of
Tylianakis(2007) on the impacts of habitat modifications to the network
structure. The data consists of 48 networks with 4090 recorded interactions. The
advantage of replicated host-parasitoid networks is that usually every
interaction is observed, not inferred from a stationary metaweb. It thus allows
to evaluate interaction probability and to factor out the effect of
co-occurrence. Five habitats were sampled along a gradient of habitat
modification: forest, abandoned coffee agroforest, coffee agroforest, pasture
and rice culture. The metaweb consists of 9 parasitoids and kleptoparasites
(Hymenoptera: Eulophidae, Ichneumonidae, Leucospidae, Megachilidae and
Chrysididae; Dyptera: Bombyliidae) of 33 species of bees and wasps (Hymenoptera:
Apidae, Megachilidae, Mutilidae, Pompilidae, Sphecidae, Vespidae). The metaweb
is illustrated at Fig. 2, along with an example of one iteration of the metaweb.

Tylianakis (2007) investigated whether habitat modification affects
the structure of these networks They found a significant impact of
the habitat on their structure, despite little variation in species
richness. Increasing habitat modification led to a higher parasitoid
to host species ratio and a paraistoid were also more specialized, thus
impacting considerably vulnerability. A closer inspection of the networks
revealed that intensive agricultural systems were dominated by a strong
interaction and a specialization of the most abundant parasitoid. Although
the discussion made clear that both the turnover in species composition and
the interaction probability changed with habitat modification, it was not
possible to partition these components.

We developed the \texttt{probaweb} package for R (REF) to fit
alternative formulations of the metaweb and the co-occurrence matrix
along an environmental gradient and run it to re-interpret the data of
Tylianaks (2007). The package provides a general interface facilitating
the development of different species and link distribution models. It is
also built to facilitate the interaction with the\texttt{mangal} database
of ecological interactions (REF). The first step consists of fitting a
probabilistic model from the observation of a pairwise interaction (binary)
and the environment (could be categorical or continuous) from the subset of
the data where the two species are co-occurring. In other words, it fits
the equation $P(L_{ijy}|X_{iy},X{jy},E_y)$ to the data where $X_{iy} = 1$
and $X_{iy} = 1$. Logistic regression was used and is currently programmed,
but alternative models could be used as well. The second steps consists of
fitting a probabilistic model for co-occurrence over the whole dataset,
$P(X_{iy},X{jy}|E_y)$, independently of the observation of an interaction. The
two probabilities are then multiplied to obtain the probability of observing
an interaction (Eq. 2). We used this probability to compute the likelihood of
each observation ($\zeta(\theta|D) = P(L_{ijy},X_{iy},X_{jy})$ if $L_{ijy}=1$
and $\zeta(\theta|D) = 1 - P(L_{ijy},X_{iy},X_{jy})$ otherwise). We then
after compare the models by their AIC.

We considered the gradient of habitat modification as a ordered categorical
variable and compared XX models (results are summarized at Table 2). Not
surprisingly the best model takes into account the effect of the environment on
both the metaweb and co-occurrence. What is most interesting are the comparisons
to the best model. First, we find that using a constant metaweb has a dramatic
impact on the fit of the model to the data (the AIC drops from X for model 1 to
X for model 2), indicating a strong effect of the environment on pairwise
interactions. Secondly, we find that the deterministic metaweb is the worst
model (model 3, AIC = ). This result indicate that the traditional approach to
consider that species interact as soon as they co-occur is definitely wrong.
Thirdly, we also find that using a constant co-occurrence does have a
significant impact on the model (the AIC drops to X, model 4), indicating there
is a non-random change in community composition with habitat modification. Taken
together, these two results better explain why network structured changed with
habitat modification, even though here we only used binary information about the
network structure. Another interesting result is that considering a neutral
co-occurrence did not impact much the fit of the model. The AIC drops to XX
with model 6, indicating that considering independent SDMs yields similar
networks over this environmental gradient. This means that for this particular
dataset, ecological interactions does not have a strong impact on species
distribution since; a strong dependence of parasitoids to the host for instance
would have a occurrence probability higher than expected by chance, while a
repulsion would have had the opposite.

@@

It is also possible to obtain for each pairwise interaction an estimate of
the uncertainty. Not surprisingly, the confidence interval is usually high
for the estimation of a probability with a small sample size. The standard
error on the evaluation of the interaction probability is provided along
with the metaweb at Fig. 3. It reveals that the uncertainty is high for
most interactions, even if 48 networks were sampled. Such an approach could
be used to detect which pairwise interaction requires additional sampling
in order to reduce the uncertainty to a manageable level.
